---
title: "Clustering with K-means, DBSCAN, and Hierarchical R Version"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

## Learning Objectives

By the end of this section, you will be able to:

- Apply K-means, DBSCAN, and Hierarchical clustering.
- Interpret common graphs.
- Produce parallel coordinates plots for each clustering solution.
- Compare cluster assignments to true Iris species.

---

## Setup

```{r}
library(tidyverse)
library(factoextra)
library(cluster)
library(dbscan)
library(NbClust)
library(GGally)   # for parallel coordinate plots
set.seed(4321)
```

---

## Dataset

```{r}
data(iris)
iris_num <- iris %>% select(where(is.numeric))
iris_scaled <- scale(iris_num)
head(iris_scaled)
```

### Explanation  
We extract the **numeric features** of the Iris dataset and scale them so that all variables contribute equally. Scaling is crucial because distance-based clustering is sensitive to variable magnitude.

---

## PCA Visualization

```{r}
pca <- prcomp(iris_scaled)
fviz_pca_ind(pca, geom="point", title="PCA Visualization of Iris")
```

### Explanation  
A PCA scatterplot allows us to see the **structure of high-dimensional data** in two dimensions. If clusters exist naturally, PCA often reveals them visually.

---

## PART 1 — Determining k for K-means

### Elbow Plot

```{r}
fviz_nbclust(iris_scaled, kmeans, method="wss") +
  labs(title="Elbow Method (WSS vs k)")
```

#### Explanation  
The elbow plot shows how total within-cluster variation decreases as k increases.  
We select the k where improvement slows down — the “elbow.”

---

### Silhouette Plot

```{r}
fviz_nbclust(iris_scaled, kmeans, method="silhouette") +
  labs(title="Silhouette Method")
```

#### Explanation  
The silhouette method evaluates how well-separated clusters are.  
Higher values indicate clearer boundaries. The optimal k maximizes silhouette width.

---

### Gap Statistic

```{r}
set.seed(4321)
gap <- clusGap(iris_scaled, FUN=kmeans, nstart=25, K.max=10, B=30)
fviz_gap_stat(gap)
```

#### Explanation  
The gap statistic compares cluster dispersion to what would be expected if no clustering existed.  
A large "gap" indicates strong underlying cluster structure.

---

### NBClust

```{r}
set.seed(4321)
nb_res <- NbClust(iris_scaled, distance="euclidean",
                  min.nc=2, max.nc=10, method="kmeans")

best_k <- nb_res$Best.nc[1]
best_k
```

```{r}
vote_dist <- as.data.frame(table(nb_res$Best.nc[1, ]))

ggplot(vote_dist, aes(Var1, Freq)) +
  geom_col(fill="steelblue") +
  labs(title="NbClust: Votes for k")
```

#### Explanation  
NBClust computes 30+ indices to recommend the number of clusters.  
We visualize the vote distribution and use the **majority vote**.

---

## K-means Clustering

```{r}
kmeans_fit <- kmeans(iris_scaled, centers = best_k, nstart=30)
fviz_cluster(kmeans_fit, iris_scaled,
             main=paste("K-means Clusters (k=", best_k, ")", sep=""))
```

### Explanation  
K-means partitions data into **spherical** clusters using centroids.  
The PCA visualization shows how those clusters look in reduced space.

---

## Compare K-means to True Species

```{r}
table(KMeans = kmeans_fit$cluster, Species = iris$Species)
```

### Explanation  
This table shows how well K-means clusters match the true species.  
K-means typically identifies Setosa perfectly but overlaps Versicolor and Virginica.

---

## Parallel Coordinates for K-means

```{r}
iris_km <- iris_num %>% mutate(Cluster = factor(kmeans_fit$cluster))
GGally::ggparcoord(iris_km, columns = 1:4, groupColumn = "Cluster",
                   title="Parallel Coordinates: K-means Clusters")
```

### Explanation  
Parallel coordinate plots show **variable profiles** across clusters.  
K-means clusters should show distinct shape patterns if clustering is meaningful.

---

## PART 2 — Hierarchical Clustering

### Dendrogram

```{r}
d <- dist(iris_scaled)
hc_complete <- hclust(d, method="complete")
fviz_dend(hc_complete, k=3, rect=TRUE,
          main="Hierarchical Clustering (Complete Linkage)")
```

#### Explanation  
Hierarchical clustering builds a **tree structure** showing merging of points/clusters.  
Cutting the dendrogram at height = 3 yields three clusters.

---

### Cut into 3 Clusters and Compare

```{r}
hc_clusters <- cutree(hc_complete, k=3)
table(Hierarchical = hc_clusters, Species = iris$Species)
```

#### Explanation  
Hierarchical clustering does not require pre-specifying k — we “cut” the dendrogram.  
This table shows how hierarchical groupings align with real species.

---

### PCA Visualization of Hierarchical Clusters

```{r}
fviz_cluster(list(data=iris_scaled, cluster=hc_clusters),
             geom="point",
             main="Hierarchical Clusters (k=3)")
```

#### Explanation  
A PCA scatterplot shows how hierarchical clusters spread in low-dimensional space.  
Complete linkage tends to produce compact clusters.

---

### Parallel Coordinates for Hierarchical

```{r}
iris_hc <- iris_num %>% mutate(Cluster = factor(hc_clusters))
GGally::ggparcoord(iris_hc, columns = 1:4, groupColumn = "Cluster",
                   title="Parallel Coordinates: Hierarchical Clusters")
```

### Explanation  
Parallel coordinates allow us to compare variable patterns across hierarchical clusters.

---

## PART 3 — DBSCAN

### kNN Distance Plot

```{r}
minPts <- 2 * ncol(iris_scaled)
kNNdistplot(iris_scaled, k=minPts)
abline(h=0.9, lty=2)
```

#### Explanation  
The kNN plot helps choose eps by showing the distance to each point’s k-th neighbor.  
The “knee” in the curve is the suggested eps threshold.

---

### Fit DBSCAN

```{r}
db_fit <- dbscan(iris_scaled, eps=0.9, minPts=minPts)
db_fit
```

#### Explanation  
DBSCAN groups high-density regions and labels low-density points as **noise (cluster 0)**.

---

### DBSCAN Clusters in PCA Space

```{r}
fviz_cluster(list(data=iris_scaled, cluster=db_fit$cluster),
             geom="point",
             ellipse.type="convex",
             main="DBSCAN Clusters")
```

#### Explanation  
DBSCAN finds **arbitrary-shaped clusters** that density-based methods capture better than K-means.

---

### Compare DBSCAN to True Species

```{r}
table(DBSCAN = db_fit$cluster, Species = iris$Species)
```

#### Explanation  
DBSCAN often segments Setosa well but may merge Versicolor and Virginica depending on eps.

---

### Parallel Coordinates for DBSCAN

```{r}
iris_db <- iris_num %>% mutate(Cluster = factor(db_fit$cluster))
GGally::ggparcoord(iris_db, columns = 1:4, groupColumn = "Cluster",
                   title="Parallel Coordinates: DBSCAN Clusters")
```

#### Explanation  
Parallel coordinates show how DBSCAN clusters differ in variable space and highlight noise points.

---

---


