---
title: "Isolation Forest - R Version"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

## Learning Objectives

After this notebook, you will be able to:

-   Understand Isolation Forest intuition\

-   Generate synthetic data with injected anomalies

-   Fit Isolation Forest

## Setup

```{r}
library(tidyverse)
library(isotree)
set.seed(4321)
```

------------------------------------------------------------------------

### 1. Isolation Forest Overview

Isolation Forest finds anomalies by isolating points with random partitions:

-   Normal points require **many** splits\
-   Outliers isolate **quickly**\
-   Fewer splits → **higher anomaly score**

------------------------------------------------------------------------

### 2. Generate Synthetic Data (Two Clusters + Outliers)

```{r}
n <- 500

cluster1 <- tibble(
  x = rnorm(n/2, 0, 1),
  y = rnorm(n/2, 0, 1),
  true_type = "normal"
)

cluster2 <- tibble(
  x = rnorm(n/2, 3, 1),
  y = rnorm(n/2, 3, 1),
  true_type = "normal"
)

# Inject anomalies
anoms <- tibble(
  x = rnorm(20, 8, 0.4),
  y = rnorm(20, -2, 0.4),
  true_type = "anomaly"
)

df <- bind_rows(cluster1, cluster2, anoms)

ggplot(df, aes(x, y, color = true_type)) +
  geom_point(size = 2) +
  labs(title = "Synthetic Data with Injected Anomalies",
       color = "True Label")
```

------------------------------------------------------------------------

### 3. Fit Isolation Forest

```{r}
iso_model <- isolation.forest(df[, c("x", "y")], ntrees = 500)

df$score <- predict(iso_model,
                    df[, c("x", "y")],
                    type = "score")

summary(df$score)
```

------------------------------------------------------------------------

### 4. Visualize Anomaly Scores

```{r}
ggplot(df, aes(x, y, color = score)) +
  geom_point(size = 2) +
  scale_color_viridis_c() +
  labs(title = "Isolation Forest Anomaly Scores",
       color = "Score")
```

------------------------------------------------------------------------

### 5. Flag Outliers (Top 5%)

```{r}
cutoff <- quantile(df$score, 0.95)
df$flag <- ifelse(df$score >= cutoff, "Outlier", "Normal")

table(df$flag)
```

```{r}
ggplot(df, aes(x, y, color = flag)) +
  geom_point(size = 2, alpha = 0.9) +
  labs(title = "Flagged Outliers (Top 5% by Score)",
       color = "Flag")
```

------------------------------------------------------------------------

### 6. Print Top Anomalies

```{r}
df %>%
  arrange(desc(score)) %>%
  slice(1:10)
```

------------------------------------------------------------------------

### 7. Benchmarked Accuracy (for synthetic data)

```{r}
table(Predicted = df$flag, Truth = df$true_type)
```

------------------------------------------------------------------------

### 8. Parallel Coordinates

This version is **100% Quarto-safe**, with no `row_number()` inside `aes()`\
and no ggparcoord dependency.

```{r}
# Clean factor
df_pc <- df %>%
  mutate(flag = factor(flag, levels = c("Normal", "Outlier")))

# Create explicit, stable row ID
df_pc <- df_pc %>%
  mutate(id = row_number())

# Long format for parallel-style plot
df_long <- df_pc %>%
  pivot_longer(cols = c(x, y),
               names_to = "Variable",
               values_to = "Value")

# Plot
ggplot(df_long,
       aes(x = Variable,
           y = Value,
           group = id,
           color = flag)) +
  geom_line(alpha = 0.5) +
  labs(title = "Parallel-Style Coordinates: Normal vs Outlier Profiles",
       color = "Flag")
```

### Interpretation

This shows how outliers differ from normal points across variables.\
If outliers consistently sit higher/lower on certain dimensions, they appear as distinct paths.

------------------------------------------------------------------------

------------------------------------------------------------------------

## PART 2 — Business-Context Example

## **Simulated Credit Transactions Dataset (Multidimensional)**

We simulate **fraud-like behavior** with:

-   Unusually large transaction amounts\
-   Transactions at unusual times\
-   Rare merchant categories\
-   Abnormal distances between transactions\
-   Too-many transactions in a short window\
-   High-risk device scores

------------------------------------------------------------------------

### 1. Generate Synthetic Credit Transaction Data

```{r}
set.seed(1234)

n <- 2000

# Normal transactions
norm <- tibble(
  amount = rgamma(n, shape=2, scale=40),
  hour = sample(6:22, n, replace=TRUE),
  merchant_risk = rbeta(n, 2, 8),
  device_score = rnorm(n, mean=50, sd=10),
  distance_km = rgamma(n, shape=2, scale=3),
  trans_per_24h = rpois(n, lambda=3),
  true_type = "normal"
)

# Fraudulent transactions
fraud <- tibble(
  amount = rgamma(60, shape=9, scale=60),             # large
  hour = sample(c(0,1,2,3,4), 60, replace=TRUE),      # late night
  merchant_risk = rbeta(60, 8, 1),                    # high risk
  device_score = rnorm(60, mean=20, sd=5),            # bad device score
  distance_km = rgamma(60, shape=1, scale=20),        # far distances
  trans_per_24h = rpois(60, lambda=12),               # high velocity
  true_type = "fraud"
)

df <- bind_rows(norm, fraud)
```

------------------------------------------------------------------------

### 2. Fit Isolation Forest

```{r}
iso <- isolation.forest(df %>% select(-true_type), ntrees=600)

df$score <- predict(iso, df %>% select(-true_type), type="score")

summary(df$score)
```

------------------------------------------------------------------------

### 3. Visualize Score Distribution

```{r}
ggplot(df, aes(score)) +
  geom_histogram(bins=40, fill="steelblue", alpha=0.8) +
  labs(title="Anomaly Score Distribution (Credit Transactions)")
```

------------------------------------------------------------------------

### 4. Flag Top 5% as Suspected Fraud

```{r}
cutoff <- quantile(df$score, 0.95)
df$flag <- ifelse(df$score >= cutoff, "Suspected Fraud", "Legit")

table(df$flag)
```

------------------------------------------------------------------------

### 5. Compare Predictions to True Fraud Labels

```{r}
table(Predicted=df$flag, Truth=df$true_type)
```

This gives insight into:

-   **True positives** (fraud detected)
-   **False positives** (legit transactions flagged)
-   **False negatives** (missed fraud)

------------------------------------------------------------------------

### 6. Visual Diagnostics

### Scatterplot: Amount vs Hour

```{r}
ggplot(df, aes(amount, hour, color=score)) +
  geom_point(alpha=0.7) +
  scale_color_viridis_c() +
  labs(title="Amount vs Hour with Anomaly Scores")
```

Fraud transactions cluster at:

-   Large amounts\
-   Extreme nighttime hours (0–4 AM)

------------------------------------------------------------------------

### Scatterplot: Merchant Risk vs Device Score

```{r}
ggplot(df, aes(merchant_risk, device_score, color=score)) +
  geom_point(alpha=0.7) +
  scale_color_viridis_c() +
  labs(title="Risky Merchant vs Bad Device Score (Anomaly Colored)")
```

Fraud behaviors:

-   High merchant risk\
-   Low device trust score

------------------------------------------------------------------------

### 7. Parallel Coordinates 

```{r}
df_pc <- df %>%
  mutate(flag = factor(flag, levels=c("Legit","Suspected Fraud"))) %>%
  mutate(id = row_number())

df_long <- df_pc %>%
  pivot_longer(cols = c(amount, hour, merchant_risk, device_score, distance_km, trans_per_24h),
               names_to = "Variable",
               values_to = "Value")

ggplot(df_long, aes(x=Variable, y=Value, group=id, color=flag)) +
  geom_line(alpha=0.15) +
  labs(title="Parallel-Style Coordinates: Fraud vs Legit",
       color="Flag")
```

This reveals fraud profiles across dimensions:

-   Higher amount\
-   Higher merchant risk\
-   Out-of-bounds distance\
-   High transaction velocity\
-   Low device reputation

------------------------------------------------------------------------

### 8. Print Most Suspicious Transactions

```{r}
df %>%
  arrange(desc(score)) %>%
  slice(1:15)
```

These represent the most abnormal behavior across multiple dimensions.

------------------------------------------------------------------------

## Summary & Business Interpretation

Isolation Forest in fraud analytics:

-   Flags **behavior that deviates from customer norms**\
-   Useful for pre-screening before rule-based or analyst review\
-   Helps detect evolving fraud patterns not captured by rules\
-   Imbalanced classes do not harm the model\
-   Supports many dimensions (amount, risk, device, geography)


------------------------------------------------------------------------
