
---
title: "Model-Agnostic Interpretability - R Version"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup-packages
#| include: true

# Core data wrangling and modeling
library(tidyverse)
library(AmesHousing)
library(ranger)
library(xgboost)

# Model-agnostic interpretability
library(iml)      # permutation importance, PD, ICE, ALE, local models, Shapley

# Alternatives / model-specific tools
library(randomForest)
```

## Learning Objectives

By the end of this section you should be able to:

- Explain what *model-agnostic* interpretability means and why it is useful.
- Compute and interpret **permutation feature importance** for several models.
- Use **partial dependence plots (PDP)** and **individual conditional expectation (ICE)** plots.
- Use **accumulated local effects (ALE)** plots and understand how they differ from PDP.
- Explain **local surrogate** ideas in the spirit of **LIME** and interpret a local explanation.
- Compute **Shapley value**–based explanations for individual predictions.
- Compare how these tools behave for:
  - a **regression** target (Sale Price), and  
  - a **binary classification** target (High vs Low price, based on the median).

---

## 1. Data and Modeling Setup

We will reuse the Ames housing data and the same subset of predictors used earlier in the book.

```{r}
#| label: load-data
#| fig-cap: "Ames housing data with selected predictors."

set.seed(4321)

ames <- make_ordinal_ames() |>
  mutate(
    id = row_number()
  ) |>
  select(
    id, Sale_Price,
    Bedroom_AbvGr,
    Year_Built,
    Mo_Sold,
    Lot_Area,
    Street,
    Central_Air,
    First_Flr_SF,
    Second_Flr_SF,
    Full_Bath,
    Half_Bath,
    Fireplaces,
    Garage_Area,
    Gr_Liv_Area, 
    TotRms_AbvGrd
  )

# Create a binary classification target: High vs Low price
median_price <- median(ames$Sale_Price)
ames <- ames |>
  mutate(
    HighPrice = if_else(Sale_Price >= median_price, "High", "Low"),
    HighPrice = factor(HighPrice, levels = c("Low", "High"))
  )

# Train / test split (same seed for reproducibility)
train_ids <- sample(ames$id, size = 0.7 * nrow(ames))
train <- ames |> filter(id %in% train_ids)
test  <- ames |> filter(!id %in% train_ids)

# Predictors only (same X for regression and classification)
X_train <- train |>
  select(-id, -Sale_Price, -HighPrice)
X_test  <- test |>
  select(-id, -Sale_Price, -HighPrice)

# Targets
y_train_reg <- train$Sale_Price
y_test_reg  <- test$Sale_Price

y_train_cls <- train$HighPrice
y_test_cls  <- test$HighPrice
y_train_cls_num <- if_else(y_train_cls == "High", 1, 0)
```

---

## 2. Models to Interpret

We will fit three models:

- A **linear regression** model (simple, transparent baseline).
- A **Random Forest** (nonlinear, interaction-rich).
- An **XGBoost** model (gradient-boosted trees).

We fit each model for the **regression** task. Then we fit classification versions for the **high/low** price target.

### 2.1 Regression Models

```{r}
#| label: fit-regression-models

# Linear regression
lm_reg <- lm(Sale_Price ~ . - id - HighPrice, data = train)

# Random Forest regression
rf_reg <- ranger(
  Sale_Price ~ . - id - HighPrice,
  data       = train,
  num.trees  = 500,
  mtry       = 4,
  importance = "permutation",
  seed       = 4321
)

# XGBoost regression: use model.matrix on the predictor set
xgb_train_reg <- model.matrix(~ ., data = X_train)

xgb_reg <- xgboost(
  data              = xgb_train_reg,
  label             = y_train_reg,
  nrounds           = 200,
  objective         = "reg:squarederror",
  max_depth         = 4,
  eta               = 0.1,
  subsample         = 0.8,
  colsample_bytree  = 0.8,
  verbose           = 0
)
```

### 2.2 Classification Models (High vs Low Price)

```{r}
#| label: fit-classification-models

# Logistic regression (glm)
glm_cls <- glm(
  HighPrice ~ . - id - Sale_Price,
  data   = train,
  family = binomial
)

# Random Forest classification
rf_cls <- ranger(
  HighPrice ~ . - id - Sale_Price,
  data        = train,
  num.trees   = 500,
  mtry        = 4,
  probability = TRUE,
  seed        = 4321
)

# XGBoost classification (probability of "High")
xgb_train_cls <- xgb_train_reg  # same predictors as regression
xgb_cls <- xgboost(
  data              = xgb_train_cls,
  label             = y_train_cls_num,
  nrounds           = 200,
  objective         = "binary:logistic",
  max_depth         = 4,
  eta               = 0.1,
  subsample         = 0.8,
  colsample_bytree  = 0.8,
  verbose           = 0
)
```

---

## 3. Wrapping Models in `iml::Predictor`

The **iml** package provides a common interface to models. Once a model is wrapped in a `Predictor`, we can apply:

- permutation importance
- PDP, ICE, ALE
- local surrogate models (LIME-style)
- Shapley value explanations

### 3.1 Regression Predictors

```{r}
#| label: iml-regression-predictors

# Predictors for regression models
pred_lm_reg <- Predictor$new(
  model = lm_reg,
  data  = X_train,
  y     = y_train_reg
)

pred_rf_reg <- Predictor$new(
  model = rf_reg,
  data  = X_train,
  y     = y_train_reg,
  predict.function = function(m, newdata) {
    predict(m, data = newdata)$predictions
  }
)

pred_xgb_reg <- Predictor$new(
  model = xgb_reg,
  data  = X_train,
  y     = y_train_reg,
  predict.function = function(m, newdata) {
    new_mat <- model.matrix(~ ., data = newdata)
    predict(m, new_mat)
  }
)
```

### 3.2 Classification Predictors (Probability of High Price)

For classification, we will interpret the **predicted probability that a home is “High” price** (above the median).

```{r}
#| label: iml-classification-predictors

# Logistic regression: probability of "High"
pred_glm_cls <- Predictor$new(
  model = glm_cls,
  data  = X_train,
  y     = y_train_cls_num,
  predict.function = function(m, newdata) {
    predict(m, newdata, type = "response")
  }
)

# Random Forest classification: use probability for "High"
pred_rf_cls <- Predictor$new(
  model = rf_cls,
  data  = X_train,
  y     = y_train_cls_num,
  predict.function = function(m, newdata) {
    preds <- predict(m, data = newdata)$predictions
    preds[, "High"]
  }
)

# XGBoost classification: probability of "High"
pred_xgb_cls <- Predictor$new(
  model = xgb_cls,
  data  = X_train,
  y     = y_train_cls_num,
  predict.function = function(m, newdata) {
    new_mat <- model.matrix(~ ., data = newdata)
    predict(m, new_mat)
  }
)
```

### 3.3 Why Some Models Need `predict.function`

The `iml` package expects the wrapped model to behave like a simple prediction function:

> **input:** a data frame of new observations  
> **output:** a numeric vector of predictions

Different R model objects behave differently:

- **`lm()`** already satisfies this contract:  
  `predict(lm_reg, newdata)` returns a numeric vector of fitted values.  
  → No custom `predict.function` is needed.

- **`ranger` (Random Forest)** returns a **list**, whose `$predictions` element contains the numeric predictions.  
  → We supply a `predict.function` that calls `predict()` and pulls out `$predictions`.

- **`xgboost`** expects a **numeric matrix**, not a data frame, and does not understand formulas or factors.  
  → We supply a `predict.function` that first converts `newdata` to a design matrix via `model.matrix()`, then calls `predict()`.

```{r}
#| label: predictor-functions-explained

# Linear model works out of the box:
# predict(lm_reg, newdata) -> numeric vector
pred_lm_reg <- Predictor$new(
  model = lm_reg,
  data  = X_train,
  y     = y_train_reg
)

# ranger: extract $predictions from the list
pred_rf_reg <- Predictor$new(
  model = rf_reg,
  data  = X_train,
  y     = y_train_reg,
  predict.function = function(m, newdata) {
    predict(m, data = newdata)$predictions
  }
)

# xgboost: convert data frame to numeric matrix
pred_xgb_reg <- Predictor$new(
  model = xgb_reg,
  data  = X_train,
  y     = y_train_reg,
  predict.function = function(m, newdata) {
    new_mat <- model.matrix(~ ., data = newdata)
    predict(m, new_mat)
  }
)
```

---

## 4. Permutation Feature Importance

**Idea:** Shuffle one feature at a time and see how much model performance degrades.  
Larger performance loss ⇒ feature was more important.

We’ll start with the **Random Forest regression** model.

```{r}
#| label: perm-importance-rf-reg
#| fig-cap: "Permutation feature importance for Random Forest regression."
#| fig-width: 7
#| fig-height: 5

set.seed(4321)
imp_rf_reg <- FeatureImp$new(
  predictor = pred_rf_reg,
  loss      = "rmse"
)

plot(imp_rf_reg)
```

You can repeat this for XGBoost:

```{r}
#| label: perm-importance-xgb-reg
#| fig-cap: "Permutation feature importance for XGBoost regression."
#| fig-width: 7
#| fig-height: 5

set.seed(4321)
imp_xgb_reg <- FeatureImp$new(
  predictor = pred_xgb_reg,
  loss      = "rmse"
)

plot(imp_xgb_reg)
```

### 4.1 Classification: Importance for Probability of High Price

Now compare to the **Random Forest classification** model, where the outcome is a *probability*.

```{r}
#| label: perm-importance-rf-cls
#| fig-cap: "Permutation feature importance for RF probability of High price."
#| fig-width: 7
#| fig-height: 5

set.seed(4321)
imp_rf_cls <- FeatureImp$new(
  predictor = pred_rf_cls,
  loss      = "ce"  # cross-entropy for probabilities
)

plot(imp_rf_cls)
```

---

## 5. Partial Dependence and ICE

**Partial Dependence Plot (PDP):**  
Average prediction as we vary one feature, holding others “as observed” and averaging over them.

**Individual Conditional Expectation (ICE):**  
One curve per observation—shows heterogeneity across individual observations.

We’ll focus on **Gr_Liv_Area** (above-ground living area) as a key driver.

### 5.1 PDP for RF Regression

```{r}
#| label: pdp-rf-reg
#| fig-cap: "Partial dependence of Sale Price on Gr_Liv_Area (Random Forest regression)."
#| fig-width: 7
#| fig-height: 5

pdp_rf_reg <- FeatureEffect$new(
  predictor = pred_rf_reg,
  feature   = "Gr_Liv_Area",
  method    = "pdp"
)

plot(pdp_rf_reg)
```

### 5.2 ICE for RF Regression

```{r}
#| label: ice-rf-reg
#| fig-cap: "ICE curves of Sale Price vs Gr_Liv_Area (Random Forest regression)."
#| fig-width: 7
#| fig-height: 5

ice_rf_reg <- FeatureEffect$new(
  predictor = pred_rf_reg,
  feature   = "Gr_Liv_Area",
  method    = "ice"
)

plot(ice_rf_reg)
```

> **Interpretation point:**  
> - PDP: one smooth curve (average effect).  
> - ICE: many curves, one per home.  
> Differences between ICE curves indicate **interactions** or **nonlinearities** that the average PDP hides.

### 5.3 Classification: PDP for Probability of High Price

Now, look at the **probability the home is High price** vs. `Gr_Liv_Area` for RF classification.

```{r}
#| label: pdp-rf-cls
#| fig-cap: "PDP of probability(HighPrice) vs Gr_Liv_Area (Random Forest classification)."
#| fig-width: 7
#| fig-height: 5

pdp_rf_cls <- FeatureEffect$new(
  predictor = pred_rf_cls,
  feature   = "Gr_Liv_Area",
  method    = "pdp"
)

plot(pdp_rf_cls)
```


---

## 6. Accumulated Local Effects (ALE)

**ALE plots** estimate the effect of a feature in a more *local* way:

- Instead of averaging over unrealistic combinations of features (a PDP problem),  
- ALE looks at changes in predictions as we move within small intervals of the feature.

We’ll compute ALE for `Gr_Liv_Area` for the RF regression model.

```{r}
#| label: ale-rf-reg
#| fig-cap: "ALE plot for Gr_Liv_Area (Random Forest regression)."
#| fig-width: 7
#| fig-height: 5

ale_rf_reg <- FeatureEffect$new(
  predictor = pred_rf_reg,
  feature   = "Gr_Liv_Area",
  method    = "ale"
)

plot(ale_rf_reg)
```

> **Teaching point:**  
> - PDP assumes features are independent; ALE relaxes this by focusing on local neighborhoods.  
> - ALE is often more reliable when predictors are correlated (as they often are in housing data).

---

## 7. Local Surrogate Models (LIME-Style)

**LIME idea:** For one specific observation:

1. Generate small perturbations around that point.  
2. Get model predictions for the perturbed points.  
3. Fit a simple, local surrogate model (e.g., a linear regression) to approximate the complex model nearby.  
4. Use the surrogate’s coefficients as a local explanation.

`iml::LocalModel` provides a LIME-style explanation.

### 7.1 Choose a House to Explain

```{r}
#| label: pick-observation

# Pick a test-set observation to explain
obs_index <- 10
x_interest <- X_test[obs_index, , drop = FALSE]
x_interest
```

### 7.2 Local Explanation for RF Regression

```{r}
#| label: lime-style-rf-reg
#| fig-cap: "Local surrogate explanation (LIME-style) for RF regression prediction."
#| fig-width: 7
#| fig-height: 5

local_rf_reg <- LocalModel$new(
  predictor   = pred_rf_reg,
  x.interest  = x_interest
)

plot(local_rf_reg)
```

### 7.3 Local Explanation for RF Classification (Probability of High Price)

```{r}
#| label: lime-style-rf-cls
#| fig-cap: "Local surrogate explanation (LIME-style) for RF classification probability of HighPrice."
#| fig-width: 7
#| fig-height: 5

local_rf_cls <- LocalModel$new(
  predictor   = pred_rf_cls,
  x.interest  = x_interest
)

plot(local_rf_cls)
```

> **Note:**  
> This is the same home, explained two different ways:
> - contribution to **predicted price**,  
> - contribution to **probability of being High price**.

---

## 8. Shapley Values

**Shapley values** come from cooperative game theory:

- Think of each feature as a “player” in a game whose payoff is the model prediction.
- Shapley values allocate the difference between the prediction and a baseline fairly across features.
- They are **local** explanations for one observation, but with nice global averaging properties.

### 8.1 Shapley Values for RF Regression

```{r}
#| label: shapley-rf-reg
#| fig-cap: "Shapley value explanation for RF regression prediction."
#| fig-width: 7
#| fig-height: 5

set.seed(4321)
shap_rf_reg <- Shapley$new(
  predictor  = pred_rf_reg,
  x.interest = x_interest
)

plot(shap_rf_reg)
```

### 8.2 Shapley Values for RF Classification

Now explain the **probability of High price** for the same home.

```{r}
#| label: shapley-rf-cls
#| fig-cap: "Shapley value explanation for RF classification probability of HighPrice."
#| fig-width: 7
#| fig-height: 5

set.seed(4321)
shap_rf_cls <- Shapley$new(
  predictor  = pred_rf_cls,
  x.interest = x_interest
)

plot(shap_rf_cls)
```

> **Interpretation:**  
> - Shapley values are additive: the sum of contributions (plus baseline) equals the prediction.  
> - Comparing regression vs classification Shapley results shows how the “story” changes when we move from dollar predictions to probabilities.

---

## 9. Alternatives and Model-Specific Tools

There are some other tools to use in R.

### 9.1 Random Forest Variable Importance with `randomForest`

Here we refit a Random Forest using the `randomForest` package and use its built-in variable importance tools.

```{r}
#| label: rf-randomForest-alt

# Create a training data frame consistent with earlier setup
training <- train |>
  select(
    Sale_Price,
    Bedroom_AbvGr,
    Year_Built,
    Mo_Sold,
    Lot_Area,
    Street,
    Central_Air,
    First_Flr_SF,
    Second_Flr_SF,
    Full_Bath,
    Half_Bath,
    Fireplaces,
    Garage_Area,
    Gr_Liv_Area, 
    TotRms_AbvGrd
  )

training.df <- as.data.frame(training)

set.seed(12345)
rf.ames <- randomForest(
  Sale_Price ~ .,
  data     = training.df,
  ntree    = 250,
  importance = TRUE
)

# Look at variable importance
varImpPlot(
  rf.ames,
  sort = TRUE,
  n.var = 10,
  main = "Top 10 - Variable Importance",
  type = 1
)

importance(rf.ames)
```

> **Comparison:**  
> - `randomForest::varImpPlot()` is *model-specific* and convenient.  
> - `iml::FeatureImp` is *model-agnostic* and can be applied to any model with a suitable prediction function.

### 9.2 Permutation Importance for Linear Regression (Again with `iml`)

```{r}
#| label: alt-linear-perm-imp

lm.ames <- lm(Sale_Price ~ ., data = training)

linear_pred <- Predictor$new(
  lm.ames,
  data = training[,-1],
  y    = training$Sale_Price,
  type = "response"
)

plot(FeatureImp$new(linear_pred, loss = "mse"))
```

### 9.3 ICE / PDP / ALE via `FeatureEffects`

The `FeatureEffects` (plural) helper in `iml` creates multiple types of effects in one object and then uses a `$plot()` method.

```{r}
#| label: alt-feature-effects

forest_pred <- Predictor$new(
  rf.ames,
  data = training[,-1],
  y    = training$Sale_Price,
  type = "response"
)

# ICE
ice_plot <- FeatureEffects$new(forest_pred, method = "ice")
ice_plot$plot("Garage_Area")

# PDP
pd_plot <- FeatureEffects$new(forest_pred, method = "pdp")
pd_plot$plot("Garage_Area")

# PDP + ICE overlay
pdice_plot <- FeatureEffects$new(forest_pred, method = "pdp+ice")
pdice_plot$plot("Mo_Sold")

# ALE
ale_plot <- FeatureEffects$new(forest_pred, method = "ale")
ale_plot$plot("Garage_Area")
```

### 9.4 LIME-Style LocalModel and Shapley (Alternative Examples)

```{r}
#| label: alt-lime-shapley

# LIME-style explanation for a particular point
point <- 132
lime.explain <- LocalModel$new(
  forest_pred,
  x.interest = training[point,-1],
  k = 5
)
plot(lime.explain)

# Shapley values for a particular point
point <- 100
shap_alt <- Shapley$new(
  forest_pred,
  x.interest = training[point,-1]
)
shap_alt$plot()
```

> **Teaching point:**  
> These examples show that:
> - You can work *entirely* within `iml` for a consistent, model-agnostic interface, or  
> - You can mix `iml` tools with model-specific utilities like `randomForest::varImpPlot()` depending on your audience and goals.

---

## 10. Summary and Comparison

Key takeaways:

- **Permutation importance** tells us which features matter *globally* for predictions.
- **PDP and ICE** show how predictions change as one feature moves:
  - PDP: average effect  
  - ICE: individual-level variation
- **ALE** is often preferred when predictors are correlated.
- **Local surrogate (LIME-style)** and **Shapley values**:
  - provide *local*, observation-level explanations,  
  - but can be averaged to create *global* stories.
- Model-specific tools (like `varImpPlot`) are convenient but tie you to a particular algorithm, whereas `iml` gives you a common language across very different models.


